Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-02 aistats/paper.tex
Some notes on stuff to cite and arguments we could make

\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}

% my packages
\usepackage{../math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size

@@

\section{Introduction and Background}
\RLP{Goal = disseminate related work to make it look nice.}
Exponential Families are an elegant and lean way to model a wide variety of data : binary, categorical, natural numbers, positive float, long or short tailed... 
They are literally the linear model of probabilities.
The exponential family for data $X\in\cX$  with sufficient statistic $T$ and natural parameter $\natp$ is the model 
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
where $\logpart$ is the log-partition function -- i.e. the normalization factor
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx 
\end{align}
where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
Note that an exponential family is entirely specified by its support set $\cX$ and its sufficient statistic $T$.
This simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.

@@

>>>>

\newcommand{\RLP}[1]{\textcolor{red}{RLP:#1}}
\newcommand{\fdk}[1]{\textcolor{Periwinkle}{fdk:#1}}
\newcommand{\TODO}[1]{\textcolor{cyan}{TODO #1}}

% my packages
\usepackage{../math_commands}
% some custom math commands
\newtheorem{proposition}{Proposition}
\newcommand*{\expect}[2][]{\ensuremath{\mathbb{E}_{#1} \left[ #2 \right] }} % expectation operator
\newcommand{\logpart}{A}
\newcommand{\conj}{\logpart^*}
\newcommand{\bregman}{\cB_\logpart}
\newcommand{\bregmanconj}{\cB_{\logpart^*}}
\newcommand{\natp}{\theta}
\newcommand{\meanp}{\mu}
\newcommand{\decrement}{D}
\newcommand{\linear}{\ell} % linearization of a function
\newcommand{\lr}{\gamma} % learning rate, or step-size
\newcommand{\lin}[1]{\left\langle#1\right\rangle}

@@

\section{Introduction and Background}

\fdk{Consider the optimization of 
\begin{equation}
	f(\theta) = \expect[x \sim p(x)]{\frac{1}{2}\norm{\theta}^2 - \lin{x, \theta}},
\end{equation}
given samples from $x$. We know how to do that one, 
and the worst-case scenario happens when $p(x)$ is a Gaussian.
Now how about the slightly more general case where
\begin{equation}
	f(\theta) = \expect[x \sim p(x)]{A(\theta) - \lin{x, \theta}}
\end{equation}
and we have full knowledge of $A$, a strictly convex twice differentiable function. 
How hard can that be?
Turns out, if $p$ is the maximum entropy distribution with 
$\expect[x \sim p(x)]{x} = \mu$, 
then it's a bit hard.}

\RLP{Goal = disseminate related work to make it look nice.}
Exponential Families are an elegant and lean way to model a wide variety of data : binary, categorical, natural numbers, positive float, long or short tailed... 
They are literally the linear model of probabilities.
The exponential family for data $X\in\cX$  with sufficient statistic $T$ and natural parameter $\natp$ is the model 
\begin{equation}
	 p(X|\natp) = \exp( \natp^\top T(X) - \logpart(\natp)) \; ,
\end{equation}
where $\logpart$ is the log-partition function -- i.e. the normalization factor
\begin{align}
    \logpart(\natp) = \log \int e^{\natp^\top T(x)} dx 
\end{align}
where the integral stands for a sum if $x$ has discrete support. \RLP{Use base measure instead.}
Note that an exponential family is entirely specified by its support set $\cX$ and its sufficient statistic $T$.
This simple model encompasses both categorical distributions $\cX = \{1, \dots, k\}$ with $T(X)$ being the one-hot encoding, and multivariate normal distributions $\cX=\real, T(X)=(X, X^2)$.

@@

\clearpage
\section{Stuff to cite and why }

\begin{itemize}
	\item Stochastic and online mirror descent: 
	We obviously have rates in this setting, but they all assume 
	that the reference function is strongly-convex relative to a norm. 
	\citep[e.g.][Thm 4.2]{bubeck2015convex}
	\item What is mirror descent called in the literature?
		\begin{itemize}
			\item Mirror Descent \citep{nemirovski1983problem,beck2003mirror} 
			\item NoLips \citep{bauschke2017descent}
		\end{itemize}
	\item Interest in mirror descent boosted by the relatively smooth view 
	of \citet{bauschke2017descent,lu2018relatively}	
	\item 
	The oroblems on how to deal with Bregman divergences abound in optimization, 
	beyond stochasticity. 
	For example, we haven't yet figured out the analog of Nesterov-type acceleration 
	on relatively smooth and strongly convex problems, 
	to bring the convergence rate from linear in $(1-\kappa)$ to $(1-\sqrt{\kappa})$, 
	or just from $1/T$ to $1/T^2$ 
	in the (non-strongly) convex case.
	\citet{dragomir2021optimal} show that naïve application of Bregman updates can not achieve acceleration.
	The tools developped to make progress on one problem might help make progress on the other. 
	\item 
	Stochastic mirror descent in the relatively smooth setting;
	\citet{hanzely2018fastest},
	\citet{dragomir2021fast}
	and the zoo of assumptions
	\item Something about the statistical asymptotic behavior for the MLE? 
	\begin{equation}
		D_A(\theta_*, \theta_n) \approx \frac{1}{2}\norm{\theta_n - \theta_*}_{\nabla^2 A(\theta_*)}^2 
		\approx \frac{1}{n}?
	\end{equation}
	\item Something about how the Bregman divergence is often not considered in the statistics lit 
	because it can be infinite and is non-symmetric and not nice to work with, 
	and so folks focus on $\expect{\norm{\theta_n - \theta_*}^2}$ even if it's somewhat meaningless 
	in the non-asymptotic regime, 
	and the best non-asymptotic results 
	still rely on Taylor expansions after $n > $ something big?
	\item Something about how we don't even know how to deal with decreasing step-sizes 
	of the order of $1/t$ 
	\item 
	Should cite Bach's line of work on SGD, averaging schemes and self-concordant functions 
	that obtain the $1/t$ rate in the smooth, convex setting instead of $1/\sqrt{t}$, 
	which is the ``expected'' rate for that setting, 
	and how this approach above is a generalization beyond the Euclidean noise model

\end{itemize}
